### 2019-06-11 2-Chapters/2-Chapter/chapter2.tex
OMG c'est 00h23 un dimanche soir et je bosse !
Commentaire Christophe Ch4.1 et 4.2

Instead of using a confidence radius $\xi_(t)$ based on Hoeffding's inequality,
the \klUCB{} algorithm rather use a specific concentration inequality for the considered exponential family, such as Chernoff-Hoeffding for Gaussian (or sub-Gaussian) rewards.
It yields improved upper confidence bounds (\ie, with smaller $\xi_(t)$) that are based on Kullback-Leibler divergence.
The indexes are computed as follows, for an exploration function $f(t)$
that is typically chosen as $f(t) \eqdef \log(t)+3\log(\log(t))$ for the theoretical analyses,
and $f(t) \eqdef \log(t)$ in practice.
The constant $c>0$ is usually chosen as $c = 1$.
\begin{align}\label{eq:klUCB_index}
    U_k^{\klUCB}(t) \eqdef \sup\limits_{q \in [0, 1]} \left\{ q : \KL\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{c f(t}{N_k(t)} \right\}.
\end{align}

@@

>>>>

Instead of using a confidence radius $\xi_k(t)$ based on Hoeffding's inequality,
the \klUCB{} algorithm rather use a specific concentration inequality for the considered exponential family, such as Chernoff-Hoeffding for Gaussian (or sub-Gaussian) rewards.
It yields improved upper confidence bounds (\ie, with smaller $\xi_k(t)$) that are based on Kullback-Leibler divergence.
The indexes are computed as follows, for an exploration function $f(t)$
that is typically chosen as $f(t) \eqdef \log(t)+3\log(\log(t))$ for the theoretical analyses,
and $f(t) \eqdef \log(t)$ in practice.
The constant $c>0$ is usually chosen as $c = 1$.
\begin{align}\label{eq:klUCB_index}
    U_k^{\klUCB}(t) \eqdef \sup\limits_{q \in [0, 1]} \left\{ q : \KL\left(\frac{X_k(t)}{N_k(t)}, q\right) \leq \frac{c f(t}{N_k(t)} \right\}.
\end{align}

@@

% % FIXME
% \newpage
% % FIXME

### 2019-06-11 2-Chapters/4-Chapter/CrownCom_17.git/IoT_slotted.tex
Commentaire Christophe Ch4.1 et 4.2

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic in decentralized IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

@@

\textbf{One gateway and many devices.}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq i$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that \emph{devices do not use sensing}, and \emph{we do not consider retransmissions of packets} (we relax this hypothesis in Section~\ref{sec:4:retransmissions}).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (two are shown in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}, in left or right).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay (typically $1\;\mathrm{s}$), to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

@@

\textbf{Static vs dynamic devices.}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, to improve their successful communication rate automatically.
\end{itemize}

@@

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the $867\;\mathrm{MHz}$ ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
Indeed, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $D \leq K$ dynamic devices.} than the number of channels}.

\textbf{Link with the MAB model:}
%
We consider the network from the point of view of one dynamic device.
Following a Bernoulli emission process, every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

The problem parameters
% are $\mu_1,\dots,\mu_K$, the mean availability of the channels,
are $S_1, \dots, S_K$ the (stationary) occupancy of the channels by the static devices,
and they are unknown to each dynamic device, so to maximize their cumulated rewards, they must learn the distributions of the channels, in order to be able to progressively improve their respective successful communication rate.
%
The goal is thus to propose a simple sequential algorithm to be applied identically and independently by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all channels a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst channels too many times.
%
Before presenting the MAB algorithms used in the experiments, we present some natural baseline (reference) policies.

@@

\paragraph{$1^{\text{st}}$ - Naive policy: Random Channel Selection} (``\textcolor{darkgreen}{Random}'')
%
We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.

@@

If the oracle chooses a fixed configuration of dynamic devices, with $D_k$ devices affected to channel $k$ is fixed (in time, \ie, stationary),
then this probability is computed as before:
%
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \Pr(\text{success}|\text{sent in channel}\;k) \; \Pr(\text{sent in channel}\;k) \nonumber \\
    & = \sum_{k=1}^{K} \underbrace{(1 - p)^{D_k - 1}}_{\;\;D_k - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_k}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_k / D }_{\;\;\text{Sent in channel}\; k\;\;}.
\end{align}

@@

We can see in Figure \ref{fig:41:from10to100} that the TS algorithm (\textcolor{red}{in red}) outperforms the \UCB{} algorithm (\textcolor{blue}{in blue}), when the number of end-devices is below 50\%. When the number of end-devices is higher, both algorithms have almost the same performance, and perform well after a small number of transmissions (\ie, they show quick convergence).
Moreover, we can see in Figures \ref{fig:41:10intelligent}, \ref{fig:41:30intelligent}, and \ref{fig:41:50intelligent} that both have better success rate than the random policy and the probability of successful transmission is between the optimal oracle and suboptimal oracle policies.
For instance, for $10\%$ of dynamic devices, after about $1000$ transmissions, using \UCB{} over the naive uniform policy improved the successful transmission rate from $83\%$ to $88\%$, and using Thompson Sampling improved it to $89\%$.
Increasing the number of end-devices decreases the gap between the optimal and random policies: the more dynamic devices, the less useful are learning algorithms, and basically for networks with only dynamic devices, the random policy is as efficient as the optimal one, as seen in Figures~\ref{fig:41:100intelligent} and on the right end side of \ref{fig:41:perf_learning}.

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.65]{perf_learning.eps}
    \caption{Learning with \UCB{} and Thomson Sampling, with many dynamic devices. Both algorithms achieve close to optimal performance after a reasonable learning time.}
    \label{fig:41:perf_learning}
\end{figure}

\textbf{Successful transmissions rate as a function of the dynamic devices.}
%
To better assess the evolution of the optimal policy compared to the random one, we have displayed on Figure~\ref{fig:41:perf_learning} the evolution of the gain, in terms of successful transmissions rate, provided by the optimal oracle and the two learning policies, after $10^6$ time slots, \ie, about $1000$ transmissions for each IoT device.
We can see that when the proportion of end-devices is low (\eg, $1\%$ of devices are dynamic), the optimal policy provides an improvement of $16\%$ compared to random channel selection.
The TS algorithm always provides near-optimal performance, but the \UCB{} algorithm has a lowest rate of convergence and performs consequently worse after $1000$ transmissions, for instance it only provides a gain of $12\%$ for the same proportion of dynamic devices ($1\%$).

@@

Figure \ref{fig:41:perf_learning} also shows that learning keeps near-optimal performance even when the proportion of devices becomes large.
Note that when this proportion increases, the assumptions of a stochastic MAB model are clearly violated, and there is no justification for the efficiency of TS and \UCB{} algorithms.
Hence, it is surprising to have near optimal performance with stochastic MAB algorithms applied to partly or fully dynamic scenarios.

>>>>

The aim of this section is to assess the potential gain of learning algorithms in IoT scenarios, even when the number of intelligent devices in the network increases,
and the network usage is more and more fluctuatating.
% and the stochastic hypothesis is more and more violated.
To do that, we suppose an IoT network made of two types of devices: \emph{static devices} that use only one channel (fixed in time), and \emph{dynamic devices} that can choose the channel for each of their transmissions. Static devices form an interfering traffic, which could be generated by devices using other standards as well.
Note that instead of assuming that each static device uses a fixed channel, we could also assume a looser hypothesis: if each static device uses a fixed sub-set of the $K$ channels, and a purely uniform random access in its set of considered channels, then \emph{in average} the observed occupancy of the $K$ channels can be modeled as if it were occupied by (more) static devices using only one channel. So our first hypothesis is actually not constraining.
We first evaluate the probability of collision if dynamic devices randomly select channels (that is, a naive approach), and if a centralized controller would optimally distribute all of them in the channels at the beginning of the scenario.
This second approach is ideal, but not realistic the most common situation of decentralized co-located IoT networks, and it is just used here as a reference.
Then, these three reference scenarios allow to evaluate the performance of bandit algorithms, such as \UCB{} and TS, in a decentralized network, in terms of successful communication rate, as it reflects the network efficiency.
We show that these algorithms have near-optimal performance, even when the proportion of end-devices increases and the interfering traffic from other devices becomes less and less stochastic.

@@

\textbf{One gateway and many devices.}
%
We consider the system model presented in Figure~\ref{fig:41:system_model1}, where a set of devices all sends uplink packets to a (unique) network gateway.
Messages can be sent in a fixed number $K\geq2$ of wireless channels,
that are assumed to be orthogonal, that is, a message in channel $k$ will not cause collision to a message in another channel $j \neq k$.
% (\eg, $K=4$ or $K=16$ for instance), which are centered indifferently around the $433.5\;\mathrm{MHz}$ or the $877\;\mathrm{MHz}$ ISM bands (as it is chosen for the implementation presented in Section~\ref{sec:4:gnuradio}).
The communication between IoT devices and this gateway is done through a simple pure acknowledged ALOHA-based protocol where devices transmit uplink packets of fixed duration whenever they want.
We mean here by ALOHA-based that \emph{devices do not use sensing}, and \emph{we do not consider retransmissions of packets} (we relax this hypothesis in Section~\ref{sec:4:retransmissions}).
%
% Note that this choice of number of channels and of frequency bands is for instance the one of the current LoRa standard in Europe, but of course the precise choice does not really matter, and future standards can choose different parameters.

@@

These communications operate in unlicensed ISM bands and, consequently, they can suffer from interference generated by uncoordinated co-localized and/or neighboring networks (two are shown in \textcolor{orange}{orange} in Figure~\ref{fig:41:system_model1}, in left or right).
This interfering traffic is uncontrolled, and is most likely unevenly distributed over the $K$ different channels,
as each IoT protocol or IoT networks can choose to use a different subset of channels.
%
In order to simulate networks designed for the IoT, we consider a protocol \textbf{with no sensing} and no repetition of uplink messages. The gateway is in charge of sending back an acknowledgement, after some fixed-time delay, to any device paired with this gateway in this network (\ie, in \textcolor{blue}{blue} in Figure~\ref{fig:41:system_model1}) who succeeded in sending an uplink packet.
%
By considering a small number of wireless channels, and a unique PHY layer configuration\footnote{~But MAB learning to also optimize on the PHY layer parameters have started to be explored, after our work \cite{Bonnefoi17}, see for instance \cite{KerkoucheAlami18}.} (\ie, modulation, waveform, etc), and in case of a non-uniform traffic in the different channels,
the device can improve their usage of the network if they are able to \emph{learn} on the fly the best channels to use, that is, the most vacant one. Indeed, in this model, the \emph{quality} of the channels (\ie, the ressources) are identified by their vacancy, or one minus their occupancy rates.

@@

\textbf{Static vs dynamic devices.}
%
We make the following assumptions on the network.
We assume that there are two types of end-devices in the network:
\begin{itemize}
    \item
    \emph{Static} end-devices are identical, and each of them uses only \emph{one} channel to communicate with the gateway, with no loss of generality.
    %  if devices random access the channels.
    Their choice is assumed to be fixed in time (\ie, stationary). The traffic generated by these devices is considered as an interfering traffic for other devices.
    \item
    \emph{Dynamic} (or \emph{smart}) end-devices can use all the available channels, by quickly changing their communication channel at any time slot, following a Machine Learning policy.
    For that end, they use their history of past communication successes or failures they experienced in each channel, to learn about channel availability.
    We also assume that the dynamic end-devices can run simple embedded decision making algorithms, and have limited but reasonable computing as well as storage capacities.
    Note that of course we assume a limited storage capacity, so no device stores the full history of communication successes and failures, but they only store average rates (which can be stored using one float number for each of the $K$ channels).
    Our goal is to propose a learning algorithm than can be implemented by each dynamic device, in a decentralized and independent manner, in order to improve their successful communication rate automatically.
\end{itemize}

@@

\textbf{Some assumptions on the network occupation:}
%
We focus on \emph{dense networks}, in which the number of devices $S + D$ is very large compared to $K$ (for instance, about $1000$ to $100000$, while $K$ is about $4$ to $256$).
As this problem is only interesting if devices are able to communicate reasonably efficiently with the gateway, we assume that devices only communicate occasionally, \ie, with a low \emph{duty cycle}, as it is always considered for IoT.
Note that even unlicensed bands have such limitations, as for instance this is a strict requirement for any device using the \SI{868}{\mega\hertz} ISM band in Europe (and its regulation also enforces a maximum transmission power, but we do not consider power transmission in this work).
%
We prefer this choice rather than non-crowded networks, \ie, where $S + D \leq K$, as the former makes more sense for realistic IoT networks.\\
% \footnote{~Here, $K$ will typically be about $10$ different radio channels, usually in the same RF band and with separate mean carrier frequencies, and the number of devices will be of the order of $1000$ to $10000$.}.
\indente
On the opposite, imagine if there were only $K=4$ channels being occupied by $D+S = 100$ devices, each communicating with a high rate of $p=1/10$, with a non-zero occupancy in each channel (\ie, $\forall k, S_k > 0$). Then, almost all time slots will lead to collisions in all channels, for a uniformly random access scheme, and thus the network efficiency (\ie, successful communication rate) will be so close to zero that using learning algorithms cannot improve much.
Such scenario is not our target of interest, and thus we prefer to only consider feasible scenarios where $p$ is smaller than $K/(S+D)$, in order to \emph{have an average of active devices not larger\footnote{~In Chapter~\ref{chapter:5} we simply consider the case of $p=1$ and $D \leq K$ dynamic devices.} than the number of channels}.

\textbf{Link with the MAB model:}
%
All devices follow a Bernoulli emission process.
Consider the network from the point of view of one dynamic device:
every time a dynamic device has to communicate with the gateway,
it has to choose one channel (at each transmission $t \geq 1, t \in \mathbb{N}$), denoted as $A(t) = k \in[K]$.
Then, the dynamic device waits in this channel $A(t)$ for an acknowledgement sent by the gateway, during a certain period (\eg, $5$ seconds for the example considered above).
Before sending another message (\ie, at time $t+1$), the dynamic device knows if it received or not this \Ack{} message.
%
For this reason, selecting channel (or arm) $k$ at time $t$ yields a (random) feedback, called a (binary) \emph{reward}, $r(t) \eqdef Y_{k,t} \in \{0,1\}$, being $0$ if no \Ack{} was received before the next message, or $1$ if \Ack{} was successfully received.
The goal of the dynamic device is to minimize its packet loss ratio, or equivalently, to maximize its number of successful transmissions, or its cumulative reward,
$\sum_{t = 1}^T r(t),$
which is the usual objective in MAB problems (see Section~\ref{sec:2:notations}).

@@

The problem parameters
% are $\mu_1,\dots,\mu_K$, the mean availability of the channels,
are $S_1, \dots, S_K$ which represent the (stationary) occupancy of the channels by the static devices,
and they are unknown to each dynamic device, so to maximize their cumulated rewards, they must learn the distributions of the channels, in order to be able to progressively improve their respective successful communication rate.
%
The goal is thus to propose a simple sequential algorithm to be applied identically and independently by each dynamic device, in a fully distributed setting (each device runs its own algorithm, from its observations), in order to minimize collisions and maximize the fraction of successful transmissions of all the dynamic devices.
%
This requires to tackle the so-called \emph{exploration-exploitation dilemma}: a device has to try all channels a sufficient number of times to get a robust estimate of their qualities, while not selecting the worst channels too many times.
%
Before presenting the MAB algorithms used in the experiments, we present some simple baseline (reference) policies.

@@

\paragraph{$1^{\text{st}}$ - Naive policy: Random Channel Selection} (``\textcolor{darkgreen}{Random}'')
%
We derive here the probability of having a successful transmission, for a dynamic device, in the case where all the dynamic devices make a purely random channel selection (\ie, uniform on $i \in [K]$).
This reflects a naive policy that could be implemented by all the dynamic devices, and it provides a reference scenario to compare against.
Note that even nowadays this is still the solution implemented for real IoT devices deployed in new LoRaWAN networks.

@@

If the oracle chooses a fixed configuration of dynamic devices, it means that for each dynamic device the oracle affects it to a unique channel for all time steps.
Then there is a number $D_k$ of devices affected to channel $k$ being fixed in time (\ie, stationary),
and thus this probability is computed as before:
%
\begin{align}\label{eq:41:prob_col}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \Pr(\text{success}|\text{sent in channel}\;k) \; \Pr(\text{sent in channel}\;k) \nonumber \\
    & = \sum_{k=1}^{K} \underbrace{(1 - p)^{D_k - 1}}_{\;\;D_k - 1 \;\text{others}\;\;} \times \underbrace{(1 - p)^{S_k}}_{\;\;\text{No static device}\;\;} \times \underbrace{ D_k / D }_{\;\;\text{Sent in channel}\; k\;\;}.
\end{align}

@@

We can see in Figure \ref{fig:41:from10to100} that the TS algorithm (\textcolor{red}{in red}) outperforms the \UCB{} algorithm (\textcolor{blue}{in blue}), when the number of end-devices is below 50\%. When the number of end-devices is higher, both algorithms have almost the same performance, and perform well after a small number of transmissions (\ie, they show quick convergence).
Moreover, we can see in Figures \ref{fig:41:10intelligent}, \ref{fig:41:30intelligent}, and \ref{fig:41:50intelligent} that both have better success rate than the random policy and the probability of successful transmission is between the optimal oracle and suboptimal oracle policies.
For instance, for $10\%$ of dynamic devices, after about $1000$ transmissions, using \UCB{} over the naive uniform policy improved the successful transmission rate from $83\%$ to $88\%$, and using Thompson Sampling improved it to $89\%$.
Increasing the number of end-devices decreases the gap between the optimal and random policies:
as expected intuitively, the more dynamic devices, the less useful are learning algorithms, and basically for networks with only dynamic devices, the random policy is as efficient as the optimal one, as seen in Figures~\ref{fig:41:100intelligent} and on the right end side of Figure~\ref{fig:41:perf_learning}.

\begin{figure}[!h]
    \centering
    \includegraphics[scale=0.65]{perf_learning.eps}
    \caption{Learning with \UCB{} and Thomson Sampling, with many dynamic devices.
        The learning gain, for each device, decreases with the proportion of dynamic devices in the network.
        Note that the values ($16\%$ etc) depend on the repartition of static devices into the $K$ channels (\ie, $S_1,\dots,S_K$) but the general profile of this plot does \emph{not} depend much on these parameters.
        % Both algorithms achieve close to optimal performance after a reasonable learning time.
    }
    \label{fig:41:perf_learning}
\end{figure}

\textbf{Successful transmissions rate as a function of the number of dynamic devices.}
%
To better assess the evolution of the optimal policy compared to the random one, we have displayed on Figure~\ref{fig:41:perf_learning} the evolution of the gain, in terms of successful transmissions rate, provided by the optimal oracle and the two learning policies, after $10^6$ time slots, \ie, about $1000$ transmissions for each IoT device.
We can see that when the proportion of end-devices is low (\eg, $1\%$ of devices are dynamic), the optimal policy provides an improvement of $16\%$ compared to random channel selection.
The TS algorithm always provides near-optimal performance, but the \UCB{} algorithm has a lowest rate of convergence and performs consequently worse after $1000$ transmissions, for instance it only provides a gain of $12\%$ for the same proportion of dynamic devices ($1\%$),
for the considered values of $S_1,\dots,S_K$.

@@

Figure \ref{fig:41:perf_learning} also shows that learning keeps near-optimal performance even when the proportion of devices becomes large.
Note that when this proportion increases, the assumptions of a stochastic MAB model are clearly violated, and there is no mathematical justification for the efficiency of TS and \UCB{} algorithms.
Hence, it is surprising to have near optimal performance with stochastic MAB algorithms applied to partly or fully dynamic scenarios.
